---
title: "Single cell RNAseq analysis of WT gastruloids (all time-points)"
format: html
editor: visual
author : Alexandre Mayran
---

```{r, setup, include=FALSE}
# choose working directory
wd <- "/data/home/ldelisle/tempAllScriptsFromMayranEtAl2023"
knitr::opts_knit$set(root.dir = wd)
```

# Basic analysis and vizualisation of WT Gastruloids

Here, we load all the required packages

```{r}
#| label: Loading packages and variable
scRNAseqFunctions.file.path <- "scRNAseq/matrices_to_plots/scRNAseqFunctions.R"
scRNAseqParam.file.path <- "scRNAseq/matrices_to_plots/scRNAseqWTparams.R"
source(scRNAseqFunctions.file.path)
source(scRNAseqParam.file.path)
```

## We define the parameters for display

```{r}
# Here you need to define the name of the folder where the RDS is
RDSfolder <- "../mountDuboule/Alex/R/RDS_new"
# This output.directory will be used for plots
output.directory <- "output.files/scRNAseq/matrices_to_plots/WT/"
# Here put name of RDS as given in step 2
nameRDS <- "combined.WT_ad2bdca4.RDS"
# Check the RDS file exists:
if(file.exists(file.path(RDSfolder, "merged", paste0(nameRDS)))){
  print(paste0(nameRDS," RDS file found, you can proceed"))
} else{
  stop(paste0(nameRDS," RDS file not found, check the location"))
}
```

## Now we load seurat object and define object specific variables: colors for time

```{r}
# Initiate the parameters which will be used for the plots
params <- list(current.fig = paste0(output.directory, "Fig.1/"), nameRDS = nameRDS, Col.featurePlot = Col.featurePlot)

if(file.exists(file.path(RDSfolder, "analyzed", paste0("analyzed.",nameRDS)))){
  print("Analyzed RDS already exist will just run the visualizations")
  combined.seurat <- readRDS(file.path(RDSfolder, "analyzed", paste0("analyzed.",nameRDS)))
  names(my.time.colors) <- levels(combined.seurat$Time)
} else{
  print("Analyzed RDS does not exist will perform analysis and visualization")
  combined.seurat <- readRDS(file.path(RDSfolder, "merged", paste0(nameRDS)))
  names(my.time.colors) <- levels(combined.seurat$Time)
  combined.seurat$TimeColor <- factor(combined.seurat$Time,
                                      levels = names(my.time.colors))
}
```

## Here we run UMAP and clustering

```{r}
#| label: UMAP and clustering
# First check the Elbow plot to determine how many PC should be used:
ElbowPlot(combined.seurat, ndims = 50)
# Then run clustering if not already present:
if(!file.exists(file.path(RDSfolder, "analyzed", paste0("analyzed.",nameRDS)))){
  combined.seurat <- RunUMAP(combined.seurat, reduction = "pca",
                             n.components = 2L,  
                             dims = 1:30, seed.use = 2)
  combined.seurat <- FindNeighbors(combined.seurat, reduction = "pca",
                                   dims = 1:30)
  combined.seurat <- FindClusters(combined.seurat, resolution = 0.85)
} else{
  print("Analyzed RDS already exist will just run the visualizations")
}

```

## Here, we identify the clusters, change their names and put the correct colors

To use this code on another dataset, please update new.cluster.name accordingly and assign them to correct identities in list.Fate.level (in scRNAseqFunctions.R)

```{r}
# choose the name you want for your cluster, order should match the 0 to n number that seurat gives.
Idents(combined.seurat) <- "seurat_clusters"
new.cluster.name <- c("Neural Tube","Som. Mes.","Post. PSM","Early NMP","Caudal Mes.",
                      "Prim. Streak","Dermomyotome","Neuron Progen. 1","Epiblast",
                      "Late NMP","Neuron Progen. 2","Sclerotome", "Ant. PSM", "ESCs",
                      "Late Endoderm", "Neuron Precursor", "Mixed Mes.", "Ant. Prim. Streak",
                      "Early Endoderm", "Pluripotent", "Unknown", "Endothelial", "Surface Ecto.")
names(new.cluster.name) <- levels(combined.seurat$seurat_clusters)
combined.seurat <- RenameIdents(combined.seurat, new.cluster.name)
# Put the new cluster name in the metadata 'Fate'
combined.seurat[["Fate"]] <- Idents(combined.seurat)
# Check all new cluster_names are in list.Fate.level
if (any(! unique(as.character(combined.seurat$Fate)) %in% unlist(list.Fate.level))) {
  stop("The following Fate are not in list.Fate.level: ", paste(unique(as.character(combined.seurat$Fate))[! unique(as.character(combined.seurat$Fate)) %in% unlist(list.Fate.level)], collapse = ", "), "\n")
} else {
  cat("all good on nomenclature")
  # Store for each cell the color in the 'FateColor' metadata
  combined.seurat[["FateColor"]] <- my.fate.colors[as.character(combined.seurat$Fate)]
  # Reorder the new cluster so the order match what was given in list.Fate.level
  combined.seurat$Fate <- factor(combined.seurat$Fate,levels =
                                   names(my.fate.colors))
}
```

We save the seurat object if needed:

```{r}
if (!file.exists(file.path(RDSfolder, "analyzed", paste0("analyzed.",nameRDS)))){
  print("Analyzed RDS does not exist will save it now...")
  dir.create(file.path(RDSfolder, "analyzed"), showWarnings = FALSE, recursive = TRUE)
  saveRDS(combined.seurat, file.path(RDSfolder, "analyzed", paste0("analyzed.",nameRDS)))
  print("Done")
} else{
  print("Analyzed RDS already exist")
}
```

## Here, we visualize the data colored by fate and by time

```{r}
#| label: Fig. 1a
# Set the directory for figure:
params$current.fig <- SwitchFigure(Fig = 1, Panel = "DimPlot.Fate.WT")
# Save a png + pdf of the plot
g <- DimPlotMayran("combined.seurat", params,
                   group.by = "Fate",
                   cols = alpha(my.fate.colors, 0.3))
# Display it
print(g) 
params$current.fig <- SwitchFigure(Fig = 1, Panel = "DimPlot.Time.WT")
g <- DimPlotMayran("combined.seurat", params,
                   group.by = "Time",
                   cols = alpha(my.time.colors, 0.3))
print(g)
```

## Display cell proportion per conditions

```{r}
#| label: Fig.1c Cluster proportion per time
params$current.fig <- SwitchFigure(Fig = "S1", Panel = "BarPlot.Fate.WT")
Idents(combined.seurat) <- "Fate"

g <- ggplot(combined.seurat[[]],
            aes(x = Time, fill = Fate)) + 
  geom_bar(position = "fill" ) +
  facet_grid(. ~ Time, scales = "free") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        # panel.grid.major = element_blank(),
        panel.grid.minor = element_blank() ) +
  scale_fill_manual(values = my.fate.colors) +
  xlab("Stage") +
  guides(fill=guide_legend(ncol=2))+
  scale_y_continuous(labels = scales::percent)
savePngPdf(g, params$current.fig,
           paste0(nameRDS,".barplot.cellFate.per.time"),
           width = 5, height = 4)
print(g)
```

## Now we highlight each individual time

```{r}
#| label: Fig.1c UMAP visualization of time Highlighted
params$current.fig <- SwitchFigure(Fig = "S1", Panel = "DimPlot.Highlight.Time.WT")
for (my.time in unique(combined.seurat$Time)){
  # Get ids of cells at the specific time
  current.cell.ids <- rownames(combined.seurat[[]][combined.seurat$Time == my.time,])
  # Highlight them
  g <- DimPlot(combined.seurat, cells.highlight = current.cell.ids,
               sizes.highlight = 0.1 ,
               pt.size = 0.1, cols.highlight = alpha("#000000",0.3)) +
    NoLegend() +
    NoAxes() +
    ggtitle(label = my.time)
  savePngPdf(g, params$current.fig,
             paste0(nameRDS, ".", my.time, ".black.highlighted"))
  print(g)
}
```

## Get markers for each Fate, display as a heatmap and save csv

Find markers and save csv:

```{r}
if (!file.exists(file.path(output.directory, "all.markers.WT.csv"))) {
  Idents(combined.seurat) <- "Fate"
  # Only use 400 cells max per Fate
  all.markers <- FindAllMarkers(combined.seurat, logfc.threshold = 0.5, min.diff.pct = 0.2, max.cells.per.ident = 400)
  write.csv(all.markers, file = file.path(output.directory, "all.markers.WT.csv"))
} else {
  print("Markers already identified")
  all.markers <- read.csv(file.path(output.directory, "all.markers.WT.csv"))
}

```

Display top 8 markers as a heatmap

```{r}
#| label: HeatMap. marker genes
params$current.fig <- SwitchFigure(Fig = "S1", Panel = "Heatmap.Fate.WT")
# Define the number of markers
nMarkers <- 8
# Select the top (ranked by avg_log2FC)
top.n <- all.markers %>% group_by(cluster) %>% top_n(n = nMarkers, wt = avg_log2FC)
# Use only 300 cells to display the heatmap
combined.seurat.small <- subset(combined.seurat, downsample = 300)
# Plot the heatmap
g <- DoHeatmap(combined.seurat.small,
               features = top.n$gene,
               size = 4, group.colors = my.fate.colors) +
  scale_fill_gradientn(colors = rev(RColorBrewer::brewer.pal(n = 10, name = "RdBu")))

savePngPdf(g, params$current.fig,
           paste0(nameRDS,".Heatmap.top", nMarkers, ".marker.per.Fate"),
           width = 27, height = 15)
print(g)
```

## Now we display the expression of desired genes

```{r}
params$current.fig <- SwitchFigure(Fig = "2", Panel = "FeaturePlot.Cdh1.Cdh2.WT")
genelist <- c("Cdh1", "Cdh2")
for(gene in genelist){
  g <- featurePlotMayran("combined.seurat", gene, params)
  print(g)
}
params$current.fig <- SwitchFigure(Fig = "S1", Panel = "FeaturePlot.Fate.marker.WT")
genelist <- c("Esrrb", "Zfp42", "Pou3f1", "Dnmt3b", # Pluripotency
              "T", "Fst", "Sp5", "Mixl1", # Transitory
              "Sox2", "Sox3", "Pax6", "Nkx6-1", # Neuronal
              "Msgn1", "Ripply2", "Meox1", "Pax1", # Mesoderm
              "Foxa2", "Sox17", "Cd34", "Wnt6") # Other
for(gene in genelist){
  g <- featurePlotMayran("combined.seurat", gene, params)
  print(g)
}
params$current.fig <- SwitchFigure(Fig = "S6", Panel = "Snai1.WT")
genelist <- c("Snai1")
for(gene in genelist){
  g <- featurePlotMayran("combined.seurat", gene, params)
  print(g)
}
```

Write sessionInfo to file:

```{r}
writeLines(capture.output(sessionInfo()), "scRNAseq/matrices_to_plots/sessionInfo_WT.txt")
```
