---
title: "Single cell RNAseq analysis of WT gastruloids (72h)"
format: html
editor: visual
author : Alexandre Mayran
---

```{r, setup, include=FALSE}
# choose working directory
wd <- "/data/home/ldelisle/tempAllScriptsFromMayranEtAl2023"
knitr::opts_knit$set(root.dir = wd)
```

# Subset only the cells of interest

## We load all the required packages

```{r}
#| label: Loading packages and variable
scRNAseqFunctions.file.path <- "scRNAseq/matrices_to_plots/scRNAseqFunctions.R"
scRNAseqParam.file.path <- "scRNAseq/matrices_to_plots/scRNAseqWTparams.R"
source(scRNAseqFunctions.file.path)
source(scRNAseqParam.file.path)
```


## We define the parameters for display

```{r}
# Here you need to define the name of the folder where the RDS is
RDSfolder <- "../mountDuboule/Alex/R/RDS_new"
# This output.directory will be used for plots
output.directory <- "output.files/scRNAseq/matrices_to_plots/WT.subset.72h/"
# Here put name of merged RDS as given in step 2
nameRDS <- "combined.WT_ad2bdca4.RDS"
# Metadata present in the combined seurat which will be used to subset:
metadata.subset <- "Sample.name"
# Values to keep
metadata.subset.values <- c("72h.WT.rep1", "72h.WT.rep2")
# Subsetted RDS
nameRDS.subset <- "subset.72h.velocity.seurat.RDS"
# To do velocyto:
LoomFolder <- "../mountDuboule/Alex/R/raw/loom/"
# Check the subsetted RDS file exists or the combined:
if (file.exists(file.path(RDSfolder, "analyzed", nameRDS.subset))){
  print(paste0(nameRDS.subset, " RDS file found, you can proceed"))
} else if(file.exists(file.path(RDSfolder, "merged", paste0(nameRDS)))){
  print(paste0(nameRDS," RDS file found, you can proceed"))
} else{
  stop(paste0(nameRDS," RDS file not found, check the location"))
}
```
## If the analyzed RDS does not exists load the big RDS and subset

### Load big RDS

```{r}
if (file.exists(file.path(RDSfolder, "analyzed", nameRDS.subset))){
  print("Analyzed RDS already exist will just run the visualizations")
  subset.seurat <- readRDS(file.path(RDSfolder, "analyzed", nameRDS.subset))
} else{
  print("Analyzed RDS does not exist will perform subset, analysis and visualization")
  combined.seurat <- readRDS(file.path(RDSfolder, "merged", paste0(nameRDS)))
}
```

### Now we subset the 72h timepoint where diversification occurs and run PCA

```{r}

if (!file.exists(file.path(RDSfolder, "analyzed", nameRDS.subset))){
  subset.seurat <- subset(combined.seurat,
                          cells = colnames(combined.seurat)
                          [combined.seurat[[]][, metadata.subset]
                            %in% metadata.subset.values])
  # We do not need combined.seurat anymore
  rm(combined.seurat)
  subset.seurat <- NormalizeData(subset.seurat,
                                 verbose = FALSE)
  subset.seurat <- FindVariableFeatures(subset.seurat,
                                        verbose = FALSE, nfeatures = 1000)
  cat("Scaling merged object..\n")
  subset.seurat <- ScaleData(subset.seurat,
                             verbose = TRUE,
                             vars.to.regress = c("percent.mt", "S.Score",
                                                 "G2M.Score"))
  manual.removal.genes <- ""
  if (length(unique(subset.seurat$Directory)) > 1) {
    cat("more than one 10x well detected, restricting variable features to limit batch effect",
        "\nvariable features will include gene with expression between 0.05 and 0.8 percentile\n")
    average.expression <- rowMeans(AverageExpression(subset.seurat, group.by = "Directory")[[1]])
    genes.to.exclude <- names(average.expression[average.expression < quantile(average.expression,
                                                                               0.05) | average.expression > quantile(average.expression,
                                                                                                                     0.8)])
    
    
    used.var.gene <- setdiff(VariableFeatures(subset.seurat),
                             c(genes.to.exclude, manual.removal.genes))
    
  } else {
    cat("Only experiments from a single 10x well is detected,\n 
        all variable genes are used as there shouldn't be batch effect\n")
    #run PCA analysis
    used.var.gene <- setdiff(VariableFeatures(subset.seurat),
                             c(manual.removal.genes))
  }
  subset.seurat <- RunPCA(subset.seurat, npcs = 50, verbose = FALSE, features = used.var.gene)
}
```

### Run UMAP and clustering

```{r}
ElbowPlot(subset.seurat, ndims = 50)

if (!file.exists(file.path(RDSfolder, "analyzed", nameRDS.subset))){
  subset.seurat <- RunUMAP(subset.seurat, reduction = "pca",
                           n.components = 2L,  
                           dims = 1:10, seed.use = 13)
  subset.seurat <- FindNeighbors(subset.seurat, reduction = "pca",
                                 dims = 1:10)
  subset.seurat <- FindClusters(subset.seurat, resolution = 0.2)
}
```

# Basic plots

## Here, we identify the clusters, change their names and put the correct colors

```{r}
Idents(subset.seurat) <- "seurat_clusters"
new.cluster.name <- c("Prim. Streak", "Ant. Prim. Streak", "Pluripotent")
names(new.cluster.name) <- levels(subset.seurat$seurat_clusters)
subset.seurat <- RenameIdents(subset.seurat, new.cluster.name)
subset.seurat[["Fate"]] <- Idents(subset.seurat)
if (any(! unique(as.character(subset.seurat$Fate)) %in% unlist(list.Fate.level))) {
  stop("The following Fate are not in list.Fate.level: ",
       paste(unique(as.character(subset.seurat$Fate))[! unique(as.character(subset.seurat$Fate)) %in% unlist(list.Fate.level)],
             collapse = ", "), "\n")
} else {
  cat("all good on nomenclature")
  subset.seurat[["FateColor"]] <- my.fate.colors[as.character(subset.seurat$Fate)]
  subset.seurat$Fate <- factor(subset.seurat$Fate,levels =
                                 names(my.fate.colors))

}
```

## Here, we visualize the data colored by fate

```{r}
# Initiate the parameters which will be used for the plots
params <- list(nameRDS = paste0("only72h_from", nameRDS), Col.featurePlot = Col.featurePlot)

params$current.fig <- SwitchFigure(Fig = "SX", Panel = "DimPlot.72h.WT")
g <- DimPlotMayran("subset.seurat", params,
                   group.by = "Fate",
                   cols = alpha(my.fate.colors, 1))
print(g)
```

## Now we display the expression of desired genes

```{r}
params$current.fig <- SwitchFigure(Fig = "1", Panel = "FeaturePlot.72h.WT")
genelist <- c("Mesp1", "Mixl1")
for(gene in genelist){
  g <- featurePlotMayran("subset.seurat", gene, params)
  print(g)
}

params$current.fig <- SwitchFigure(Fig = "S2", Panel = "FeaturePlot.72h.WT")

genelist <- c("Eomes", "Lhx1", "Otx2", "Sox17")
for(gene in genelist){
  g <- featurePlotMayran("subset.seurat", gene, params)
  print(g)
}

params$current.fig <- SwitchFigure(Fig = "S3", Panel = "FeaturePlot.72h.WT")

genelist <- c("Frzb", "Snai1", "Fst", "Pou5f1")
for(gene in genelist){
  g <- featurePlotMayran("subset.seurat", gene, params)
  print(g)
}
```

# Get correlation hints

## Here we will extract the expression of some genes of PS and APS to infer the correlation between genes of this program in single cell

```{r}
subset.seurat <- add.level(subset.seurat, "Lhx1")
subset.seurat <- add.level(subset.seurat, "Mesp1")
params$current.fig <- SwitchFigure(Fig = "S2", Panel = "VlnPlot.Split.level.72hWT")

g <- VlnPlot(subset.seurat, features = c("Mesp1", "Mixl1", "Eomes",
                                         "Lhx1", "Gsc", "Otx2",
                                         "Gata6", "Snai1", "Cdh2",
                                         "T", "Pou5f1", "Cdh1"),
             group.by = "Mesp1_level", cols = grey.colors(start = 0.9, end = 0.3, n = 4), pt.size = 0, ncol = 3)
print(g)
savePngPdf(g, params$current.fig,
           paste0(nameRDS, ".VlnPlot.", "Mesp1_level.", "sc.correl"),
           width = 8, height = 12)
g <- VlnPlot(subset.seurat, features = c("Lhx1", "Mixl1", "Eomes",
                                         "Mesp1", "Gsc", "Otx2",
                                         "Gata6", "Snai1", "Cdh2",
                                         "T", "Pou5f1", "Cdh1"),
             group.by = "Lhx1_level", cols = grey.colors(start = 0.9, end = 0.3, n = 4), pt.size = 0, ncol = 3)
print(g)
savePngPdf(g, params$current.fig,
           paste0(nameRDS,".VlnPlot.", "Lhx1_level.", "sc.correl"),
           width = 8, height = 12)

```

# Run RNA velocity

## Now let's run some RNA velocity analysis (if not already computed)

```{r}
if (!file.exists(file.path(RDSfolder, "analyzed", nameRDS.subset))){
  # The barcodes in the loom file are like: "72h_WT_rep1:AACCACAAGCAATTCCx" "72h_WT_rep1:AACAACCAGTACTCGTx" "72h_WT_rep1:AACAAGAAGTGGTTAAx"
  # While in Seurat they are like: "72h.WT.rep1_AAACCCAAGCTGGCCT" "72h.WT.rep1_AAACCCACATGACTCA" "72h.WT.rep1_AAACGAAAGCCACAAG"
  # Generate a named vector for conversion:
  subset.seurat.barcodes <- sapply(strsplit(colnames(subset.seurat), "_"), tail, 1)
  subset.seurat.barcodes.conversion <- paste0(subset.seurat$RDS, "_", subset.seurat.barcodes)
  names(subset.seurat.barcodes.conversion) <- paste0(gsub(pattern = "\\.", replacement = "_", x = subset.seurat$orig.ident ), ":", subset.seurat.barcodes, "x")
  # Make vector with loom paths
  loom.files <- file.path(LoomFolder, paste0(unique(subset.seurat$Directory), ".loom"))
  names(loom.files) <- unique(subset.seurat$Directory)
  # Read them
  matrices <- lapply(loom.files, ReadVelocity)
  # Convert to Seurat to more easily rename cells and genes and transfer assays:
  seurats <- lapply(matrices, as.Seurat)
  # Combine Seurat objects keeping the same order:
  Combined.seurat.velocity <- merge(seurats[[as.character(unique(subset.seurat$Directory))[1]]],
                                    y = subsetByNamesOrIndices(seurats,
                                                               as.character(unique(subset.seurat$Directory))[2:length(as.character(unique(subset.seurat$Directory)))]),
                                    project = "aggregate.velocity")
  # Restrict to cells which are on the subset.seurat object:
  Combined.seurat.velocity <- subset(Combined.seurat.velocity, cells = names(subset.seurat.barcodes.conversion))
  # Rename cells
  Combined.seurat.velocity <- RenameCells(Combined.seurat.velocity, new.names = unname(subset.seurat.barcodes.conversion[colnames(Combined.seurat.velocity)]))
  # Transfer assays
  subset.seurat[["spliced"]] <- Combined.seurat.velocity[["spliced"]]
  subset.seurat[["unspliced"]] <- Combined.seurat.velocity[["unspliced"]]
  # Run velocity
  subset.seurat <- RunVelocity(subset.seurat,
                               spliced.average = 0.2, unspliced.average = 0.02,
                               deltaT = 1, kCells = 20, fit.quantile = 0.02,
                               group.by = "Fate")
  # Save RDS
  dir.create(file.path(RDSfolder, "analyzed"), showWarnings = FALSE, recursive = TRUE)
  saveRDS(subset.seurat, file.path(RDSfolder, "analyzed", nameRDS.subset))
}
```

## Display the velocity on UMAP

### Compute the velocity on UMAP

```{r}
if(!file.exists(file.path(RDSfolder, "analyzed", "velocity.on.embedding.72h.RDS"))){
  emb <- Embeddings(subset.seurat, reduction = "umap")
  
  velocity.on.embedding <- show.velocity.on.embedding.cor(emb, subset.seurat@tools[["RunVelocity"]],
                                                          n = 300, scale='sqrt',
                                                          cell.colors = ac(subset.seurat$FateColor,
                                                                           alpha=0.5),cex=0.8,arrow.scale = 5,
                                                          show.grid.flow=TRUE,
                                                          min.grid.cell.mass = 0.5, grid.n=40, arrow.lwd = 1,
                                                          do.par = F, cell.border.alpha = 0, return.details = T,
                                                          n.cores = 3)
  saveRDS(velocity.on.embedding, file = file.path(RDSfolder, "analyzed", "velocity.on.embedding.72h.RDS"))
} else{
  velocity.on.embedding <- readRDS(file.path(RDSfolder, "analyzed", "velocity.on.embedding.72h.RDS"))
}
```

### Plot it

```{r}
# We use ggplot2 to display the arrows as they are more customizable
params$current.fig <- SwitchFigure(Fig = "2", Panel = "RNA.Velocity")

g <- DimPlot(subset.seurat, cols = my.fate.colors, label = F, pt.size = 1) +
  geom_segment(data = as.data.frame(velocity.on.embedding$garrows),
               aes(x = x0, y = y0, xend = x1, yend = y1),
               arrow = arrow(length = unit(1, "mm"), type = "open", angle = 20), alpha = 0.5) +
  NoLegend() + NoAxes()
print(g)

savePngPdf(g, params$current.fig,
           paste0(nameRDS, ".RNA.velocity.72h"),
           width = 6, height = 4)
```

### Show gene specific velocity

```{r}
params$current.fig <- SwitchFigure(Fig = "2", Panel = "Gene.Velocity.72h")
gene.list <- c("Cdh1", "Cdh2")

gene.list.intersected <- intersect(gene.list, rownames(velocity.on.embedding[["vel"]]))

if (length(gene.list) != length(gene.list.intersected)) {
  print(paste("the following genes do not have velocity:", paste(setdiff(gene.list, rownames(velocity.on.embedding[["vel"]])), collapse = ", ")))
}

for(gene in gene.list.intersected){
  g <- ggplot(as.data.frame(Embeddings(object = subset.seurat,reduction = "umap")), aes(x = UMAP_1, y = UMAP_2)) + 
    geom_point(aes(colour = velocity.on.embedding[["vel"]][gene,])) +
    scale_colour_gradientn(paste(gene, "velocity"),
                           colors = RColorBrewer::brewer.pal(n = 11, name = "RdBu")) +
    theme_minimal(base_line_size = 0.1) +
    NoAxes()
  print(g)
  savePngPdf(g, params$current.fig,
             paste0(gene, ".specific.Velocity",nameRDS,".72h"),
             width = 8, height = 4)
}

params$current.fig <- SwitchFigure(Fig = "S3", Panel = "Gene.Velocity.72h")
gene.list <- c("Pou5f1", "Fst")

gene.list.intersected <- intersect(gene.list, rownames(velocity.on.embedding[["vel"]]))

if (length(gene.list) != length(gene.list.intersected)) {
  print(paste("the following genes do not have velocity:", paste(setdiff(gene.list, rownames(velocity.on.embedding[["vel"]])), collapse = ", ")))
}

for(gene in gene.list.intersected){
  g <- ggplot(as.data.frame(Embeddings(object = subset.seurat,reduction = "umap")), aes(x = UMAP_1, y = UMAP_2)) + 
    geom_point(aes(colour = velocity.on.embedding[["vel"]][gene,])) +
    scale_colour_gradientn(paste(gene, "velocity"),
                           colors = RColorBrewer::brewer.pal(n = 11, name = "RdBu")) +
    theme_minimal(base_line_size = 0.1) +
    NoAxes()
  print(g)
  savePngPdf(g, params$current.fig,
             paste0(gene, ".specific.Velocity",nameRDS,".72h"),
             width = 8, height = 4)
}
```
